---
title: "COVID 19 John Hopkins"
author: "Colin H."
date: "`r Sys.Date()`"
output: pdf_document
---

# Outline

## Introduction

### Project Objective

```         
- The aim here is to analyze and describe the data provided. Furthermore, this will be a basic linear regression analysis plotting cases against deaths. 
-NOTE: For sake of readability, I have not included the descriptions for each dataset. However, I have included in a textual format how and what I included and removed. 
```

### Data Source and Description

```         
- In regards to the data, I will utilize the global dataset. TODO: However, I have maintained the different versions 
- COVID19 dataset from the Johns Hopkins Github site.
    - [https://github.com/CSSEGISandData/COVID-19](https://github.com/CSSEGISandData/COVID-19)
- This includes a complete list of all reported COVID information from the sources used in the data set, since January 21, 2020 until March 10, 2023, when Johns Hopkins ceased its collecting and reporting of global COVID-19 data.
```

<https://github.com/CSSEGISandData/COVID-19>

## Libraries

```{r, message=FALSE }
library(tidyverse)
```

## Including Datasets

```{r Datasets , message=FALSE}
url_in <-
"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
"time_series_covid19_confirmed_global.csv",
"time_series_covid19_deaths_US.csv",
"time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)
```

## Reading in Data

```{r Read, message= FALSE}

global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])
```

## Pre-Process/Tidy

There are a few columns, I will get rid of. TODO: Which columns do I not need?

```{r Preprocess, message=FALSE}
global_cases <- global_cases %>%
  pivot_longer(cols = -c("Province/State", "Country/Region", "Lat", 'Long'),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c("Province/State", "Country/Region", "Lat", 'Long'),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

#join cases with deaths
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))


global <- global %>%
  filter(cases > 0) %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

#TODO UID Lookup for Comparative analysis 
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

#TODO: encapsulate global; refactor based on 6:30 of Tidying video
global <- global %>%
  #joins Country Region to Province/State
  left_join(uid, by = c("Province_State", "Country_Region" )) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key )

```

```{r}

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               #date is chr instead of date object
               names_to = "date",
               values_to = "cases") %>%
  #slices Admin2 to cases 
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

#join US
US <- US_cases %>%
  full_join(US_deaths)


#Regularization of cases over population 
US_by_state <- US%>%
  group_by(Province_State, Country_Region, date) %>%
  #For each state, cases = sum of cases, each state had multiple counties
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  #TODO: ensuer number is right below 
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state%>%
  group_by(Country_Region, date) %>%
  #For each state, cases = sum of cases, each state had multiple coun countries
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  
  #TODO: ensuer number is right below 
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
```

## EDA & Visualization

```{r Global EDA}
global %>%
  #TODO: Determine if or not to have specific country.... Sweden? 
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "Cases"), size = 1) +
  geom_point(aes(color = "Cases"), size = 2) +
  geom_line(aes(y = deaths, color = "Deaths"), size = 1, linetype = "dashed") +
  geom_point(aes(y = deaths, color = "Deaths"), size = 2, shape = 16) +
  scale_y_log10() +
  labs(title = "Cases vs Deaths Over Time",
       x = "Date",
       y = "Count of...",
       color = "Legend") +
  scale_color_manual(name = "Legend",
                     values = c("Cases" = "blue", "Deaths" = "grey"),
                     labels = c("Cases", "Deaths")) +

  theme_minimal()

#TODO Enter new Cases
global <- global%>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

#had to be modified previously
global %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "Cases"), size = 1) +
  geom_point(aes(color = "Cases"), size = 2) +
  geom_line(aes(y = new_deaths, color = "Deaths"), size = 1, linetype = "dashed") +
  geom_point(aes(y = new_deaths, color = "Deaths"), size = 2, shape = 16) +
  scale_y_log10() +
  labs(title = "Cases vs Deaths Over Time",
       x = "Date",
       y = "Count",
       color = "Legend") +
  scale_color_manual(name = "Legend",
                     values = c("Cases" = "blue", "Deaths" = "grey"),
                     labels = c("Cases", "Deaths")) +

  theme_minimal()

max(global$date)
max(global$cases)
```

```{r eval=FALSE, include=FALSE}
state = 'Wisconsin'
US_totals %>%
  #filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "Cases"), size = 1) +
  geom_point(aes(color = "Cases"), size = 2) +
  geom_line(aes(y = deaths, color = "Deaths"), size = 1, linetype = "dashed") +
  geom_point(aes(y = deaths, color = "Deaths"), size = 2, shape = 16) +
  scale_y_log10() +
  labs(title = "Cases vs Deaths Over Time",
       x = "Date",
       y = "Count",
       color = "Legend") +
  scale_color_manual(name = "Legend",
                     values = c("Cases" = "blue", "Deaths" = "red"),
                     labels = c("Cases", "Deaths")) +

  theme_minimal()


max(US_totals$date)
max(US_totals$cases)

#TODO Enter new Cases
US_by_state <- US_by_state%>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
  
US_totals <- US_totals%>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

#had to be modified previously
US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "Cases"), size = 1) +
  geom_point(aes(color = "Cases"), size = 2) +
  geom_line(aes(y = new_deaths, color = "Deaths"), size = 1, linetype = "dashed") +
  geom_point(aes(y = new_deaths, color = "Deaths"), size = 2, shape = 16) +
  scale_y_log10() +
  labs(title = "Cases vs Deaths Over Time",
       x = "Date",
       y = "Count",
       color = "Legend") +
  scale_color_manual(name = "Legend",
                     values = c("Cases" = "blue", "Deaths" = "red"),
                     labels = c("Cases", "Deaths")) +

  theme_minimal()

tail(US_totals %>% select(new_cases, new_deaths, everything()))

US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000*cases / population,
            deaths_per_thou = 1000*deaths / population) %>%
  filter(cases > 0, population > 0)

#TODO updated Slice min/max
US_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  slice_max(deaths_per_thou, n = 10) %>%
select(deaths_per_thou, cases_per_thou, everything())

```

## Model

```{r, Global Model}
#basic GLM model
glm_model <- glm(deaths ~ cases, data = global)

summary(glm_model)

# Visualize model results
predicted_counts <- predict(glm_model, type = "response")

ggplot(global, aes(x = Combined_Key, y = cases)) +
  geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7) +
  geom_line(aes(y = predicted_counts), color = "red", size = 1) +
  labs(title = "Observed vs. Predicted COVID-19 Counts by Country", x = "Country", y = "Number of Cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r eval=FALSE, include=FALSE}

# Fit linear regression model
model <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

summary(model)


US_state_totals %>% mutate(pred = predict(model))

US_tot_w_pred  <- US_state_totals %>% mutate(pred = predict(model))
US_tot_w_pred %>% ggplot() + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

## Metrics

```{r Metrics}

```

## Communication of Results

-   Interpreting the Model Output
-   Key Insights and Findings

## Bias/Conclusion

-   Summary of Findings
    -   TODO: Explain what had happened
-   Limitations [Bias]
    -   There is Bias towards under reporting of cases
    -   However, there is potentially overeporting of deaths What gets included as a death? Is someone dying from COVid or with COVID?
    -   My bias is pointed towards how I feel about Covid
-   Further Research = TODO: What would I like to see or to know?
    -   There would be interesting

## Appendix [Session Info ]

TODO: Sources to be placed here?

```{r}
sessionInfo()
```
